{ \sffamily \LARGE Electronic supplementary material}\\{\bf \sffamily \LARGE Evolution of reciprocity with limited payoff memory

### 2024-03-15 paper/si.tex

\def\alld{\texttt{ALLD}}
\def\tft{\texttt{TFT}}
\def\gtft{\texttt{GTFT}}
\def\rolemodel{\text{RM}}
\def\learner{\text{L}}
\def\resident{\texttt{R}}
\def\mutant{\texttt{M}}
\def\strategy{s}
\def\esm{electronic supplementary material}

@@

So far, we have discussed the prisoner's dilemma. However, this is only one
example of social dilemmas. Further examples of social dilemmas include the
snowdrift game~\cite{doebeli:Ecology:2005}, the stag hunt
game~\cite{pacheco:ProcB:2009}, and the public goods
game~\cite{pinheiro:PLoSCompBio:2014}. In the following, we discuss how our
methodology can be applied to other games. More specifically, here we focus on
the snowdrift game.

Similarly to the prisoner's dilemma, the snowdrift game is a two-player game. In
each round, individuals can choose whether to cooperate ($C$) or defect ($D$).
Here, cooperation yields a benefit \(b\) to the cooperator as well as to the
opposing player and incurs a cost \(c\) if the opponent defects, but only a cost
\(c/2\) if the opponent cooperates. This results in the following payoffs:

\begin{align}
	\bordermatrix{%
		& C & D \cr
		C &\ b -c/2 &\ b-c \  \cr
		D &\ b &\ 0\ \cr
	} . 
\end{align}\label{eq:snowdrift_payoff}
If \(b>c>0\), as before, then cooperation is a better strategy than defection if the opponent
plays defection. On the other hand, if the opponent plays C, then D is still the best
response. Thus, both strategies can invade when rare, resulting in a mixed
evolutionarily stable state at which the proportion of cooperators is
\(1-c/(2b-c)\). It is important to note that in this state, the population
payoff is smaller than it would be if everybody played C; hence, the SD still
represents a social dilemma.

In the following, we discuss our methodology for the snowdrift game. We start by
considering the case of perfect memory, and then we extend our analysis to the
case of limited memory. More specifically, we look again at which conditions
cooperation is stochastically stable for the two extreme cases of perfect and
limited memory. We then present simulation results for all the six cases
explicitly considered in this work."

@@

{\bf Perfect payoff memory.} We have already discussed the general setup for the
case of perfect memory in Section~\ref{section:perfect_memory}. Here, we derive
the conditions under which cooperation is stochastically stable for the
snowdrift game. As a reminder, we suppose that there is a single mutant,
\(k\!=\!1\) that adopts the strategy $\alld$, while the residents adopt the
strategy \(\gtft\). The resulting average distributions between the residents
and the mutants remain the same as in Section\ref{section:perfect_memory}.
However, the expected payoffs of the strategies now differ, as we are now
considering the Snowdrift payoff matrix~\eqref{eq:snowdrift_payoff}.
Specifically, the expected payoffs are computed as,

@@

\begin{align*}
\frac{\lambda_1^{-}}{\lambda_1^{+}} = e^{-\beta \left( (1-\delta+\delta q)(b - \frac{b-c}{N-1})- \frac{N-2}{N-1}(b-c/2) \right)}
\end{align*}
By definition, cooperation is stochastically stable if this ratio exceeds one, which is equivalent to
\begin{equation}\label{Eq:Condition_PerfectMemory_Snowdrift}
q<1-\frac{1}{\delta}\cdot \frac{N(c/2)}{(N\!-\!2)b+c}.
\end{equation}
For such a strategy to be feasible we require $q\!>\!0$, which implies $\delta
\!>\! \big((N\!-\!2)b+c\big)/\big(N(c/2)\big)$. In particular, in the limit of
large populations \(N \!\rightarrow\! \infty\), we obtain that cooperation is
stochastically stable if  \(q \!<\! 1\! -\! c/(2\delta b)\).

@@

\begin{equation*}
\frac{\lambda^{-}_1}{\lambda^{+}_1} = \frac{ \frac{N - 2}{N - 1}  
\left(\frac{ 1 - \delta + \delta q}{1+e^{ \beta (c/2)} } 
  + \frac{ \delta  \left(1 - q \right)}{1+ e^{-  \beta  \left(b - c/2 \right)}} \right)  
  +  \frac{1}{N-1}  \left(\frac{1 - \delta + \delta q}{1 + e^{\beta c}}
  + \frac{ \delta  \left(1 - q \right)}{2} \right)}
  { \frac{N - 2}{N - 1}  
\left( \frac{ 1 - \delta + \delta q}{1 + e^{-\beta (c/2)}}
+  \frac{ \delta  (1 - q)}{1 +e^{  \beta (b - c/2)}} \right) 
+  \frac{1}{N -1} \left(\frac{ 1 - \delta + \delta q}{1 + e^{-  \beta c}}
+  \frac{ \delta  \left(1 - q \right)}{2} \right)} .
\end{equation*}
Cooperation is stochastically stable if $\lambda^-_1/\lambda^+_1\!>\!1$. 
While one can solve this inequality for $q$, the resulting condition is somewhat lengthy.  
To obtain a more interpretable condition, we consider the limit of strong selection \(\beta\! \rightarrow\! \infty\) and
large populations \(N \!\rightarrow \!\infty \). In that case, because $b\!>\!c\!>\!0$, the above ratio simplifies to
\begin{equation}
    \frac{\lambda^{-}_1}{\lambda^{+}_1} = \frac{\delta(1 - q)}{1 - \delta + \delta q}.
\end{equation}

This condition is identical to condition~\eqref{Eq:TransitionRatioSimple} for
the prisoner's dilemma.

@@

@@

{
{\setlength{\bibsep}{0\baselineskip}
\bibliographystyle{naturemag}
\bibliography{bibliography}
}

>>>>

\def\alld{\texttt{ALLD}}
\def\allc{\texttt{ALLC}}
\def\tft{\texttt{TFT}}
\def\gtft{\texttt{GTFT}}
\def\rolemodel{\text{RM}}
\def\learner{\text{L}}
\def\resident{\texttt{R}}
\def\mutant{\texttt{M}}
\def\strategy{s}
\def\esm{electronic supplementary material}

@@

So far, we have discussed the repeated donation game, which is only one example
of social dilemmas. In the following section, we explore how our methodology can
be applied to other games, with a focus on the snowdrift game.

Similar to the donation game, the snowdrift game is a two-player game where
individuals choose to cooperate (\(C\)) or defect (\(D\)) each round.
Cooperation yields a benefit \(b\) to both players, incurring a cost \(c\) if
the opponent defects, but only \(c/2\) if the opponent cooperates. This results
in the following payoff matrix

\begin{align}
	\bordermatrix{%
		& C & D \cr
		C & b - \frac{c}{2} & b - c  \cr
		D & b & 0 \cr
	} . 
\end{align}\label{eq:snowdrift_payoff}

In the following subsections, we discuss how our results extend to the snowdrift
game. In Section~\ref{section:stochastic_stability_snowdrift}, we investigate
under which conditions cooperation is stochastically stable for both perfect and
limited memory scenarios. In Section~\ref{section:simulation_results_snowdrift},
we present simulation results for all six cases considered in this study.

@@

{\bf Perfect payoff memory.} We have already discussed the general setup for the
case of perfect memory in Section~\ref{section:perfect_memory}. Here, we derive
the conditions under which cooperation is stochastically stable for the
snowdrift game. As a reminder, we suppose that there is a single mutant,
\(k\!=\!1\) that adopts the strategy $\alld$, while the residents adopt the
strategy \(\gtft\). The resulting average distributions between the residents
and the mutants remain the same as in Section\ref{section:perfect_memory}.
However, the expected payoffs of the strategies now differ, as we are now
considering the snowdrift payoff matrix~\eqref{eq:snowdrift_payoff}. \\

\noindent
Specifically, the expected payoffs are computed as,

@@

\begin{align*}
\frac{\lambda_1^{-}}{\lambda_1^{+}} = e^{-\beta \left( (1-\delta+\delta q)(b - \frac{b-c}{N-1})- \frac{N-2}{N-1}(b-c/2) \right)}
\end{align*}
By definition, cooperation is stochastically stable if this ratio exceeds one, which is equivalent to
\begin{equation}\label{Eq:Condition_PerfectMemory_Snowdrift}
q<1-\frac{1}{\delta}\cdot \frac{N(c/2)}{(N\!-\!2)b+c}.
\end{equation}
For such a strategy to be feasible we require $q\!>\!0$, which implies $\delta
\!>\! \big((N\!-\!2)b+c\big)/\big(N(c/2)\big)$. In particular, in the limit of
large populations \(N \!\rightarrow\! \infty\), we obtain that cooperation is
stochastically stable if  \(q \!<\! 1\! -\! c/(2\delta b)\). \\

@@

\begin{equation*}
\frac{\lambda^{-}_1}{\lambda^{+}_1} = \frac{ \frac{N - 2}{N - 1}  
\left(\frac{ 1 - \delta + \delta q}{1+e^{ \beta (c/2)} } 
  + \frac{ \delta  \left(1 - q \right)}{1+ e^{-  \beta  \left(b - c/2 \right)}} \right)  
  +  \frac{1}{N-1}  \left(\frac{1 - \delta + \delta q}{1 + e^{\beta c}}
  + \frac{ \delta  \left(1 - q \right)}{2} \right)}
  { \frac{N - 2}{N - 1}  
\left( \frac{ 1 - \delta + \delta q}{1 + e^{-\beta (c/2)}}
+  \frac{ \delta  (1 - q)}{1 +e^{  \beta (b - c/2)}} \right) 
+  \frac{1}{N -1} \left(\frac{ 1 - \delta + \delta q}{1 + e^{-  \beta c}}
+  \frac{ \delta  \left(1 - q \right)}{2} \right)} .
\end{equation*}
Cooperation is stochastically stable if $\lambda^-_1/\lambda^+_1\!>\!1$. 
We consider the limit of strong selection \(\beta\! \rightarrow\! \infty\) and
large populations \(N \!\rightarrow \!\infty \). In that case, because $b\!>\!c\!>\!0$, the above ratio simplifies to
\begin{equation}\label{Eq:Condition_LimitedMemory_Snowdrift}
    \frac{\lambda^{-}_1}{\lambda^{+}_1} = \frac{\delta(1 - q)}{1 - \delta + \delta q}.
\end{equation}

Interestingly, this condition is identical to
condition~\eqref{Eq:TransitionRatioSimple} for the donation game. To understand
this result, we first need to discuss the difference between the two games. The
difference between the two games lies in the relationship between the possible
payoffs. More specifically, in the donation game, the payoffs satisfy the
relationship \(T > R > P > S\), whereas in the snowdrift game, it's \(T > R > S
> P\).
Now, note that for the above condition, the relationship between all the
possible payoffs is not important; what is important is the relationship between
all the feasible pairs of payoffs. In a population of \gtft{} strategies with a
single mutant of \alld{}, the feasible pairs of payoffs are \((R, T)\), \((R,
P)\), \((S, T)\), and \((P, P)\). In both games, the payoff \(T\) is higher than
\(S\) and \(R\), and \(R\) is higher than the punishment payoff, thus the
condition remains unchanged.

@@

To investigate whether the previous static observations accurately depict the
dynamics of evolving populations, we once again rely on simulations. Firstly, we
conducted separate simulations for perfect and limited payoff memory. In each
scenario, we examined both low and high benefits of cooperation (\(b/c=2\) and
\(b/c=10\), respectively). Supplementary Fig.
\ref{fig:expected_and_stochastic_for_snowdrift} illustrates the conditional
cooperation probabilities \(p\) and \(q\) of the evolving populations. Overall,
we find that the evolving populations are more cooperative compared to those of
the donation game. There is no cluster around the neighborhood of \alld{} as
they are easily outperformed by more cooperative strategies. Moreover, the
evolving cooperation rates are high for all the cases. Even in the limited
payoff memory for low benefit, the evolving cooperation is at 71\%. Thus, the
in the snowdrift game, the memory requirement for cooperation is
even more modest compared to the donation game.

\begin{figure}[t]
  \centering
  \includegraphics[width=.75\textwidth]{static/snowdrift_game_results.pdf}
  \caption{{\bf Evolutionary dynamics under perfect and limited payoff memory.}
The four panels depict the outcome of evolutionary simulations for the repeated
snowdrift game with either a low or high benefit of cooperation. In the top
panels, individuals have perfect payoff memory, meaning they consider all their
past interactions against all population members, taking every turn of each
repeated game into account. In the bottom panels, individuals have limited
payoff memory, considering only their very last interaction against one specific
population member, and taking into account only one round of the repeated game.
Colors represent how often the respective region of the strategy space is
visited over time.
In all cases, there is one dominant cluster, and that is the cluster of
conditionally cooperative strategies with \(p \approx 1\), where \(q\) satisfies
the constraints \eqref{Eq:Condition_PerfectMemory_Snowdrift} and
\eqref{Eq:Condition_LimitedMemory_Snowdrift}, respectively.
Simulations are run for \(T = 10^7\) time steps, using a cost \(c = 1\), a
continuation probability of \(\delta = 0.999\), and a selection strength of
\(\beta = 1\), in a population of size \(N = 100\).}
\label{fig:expected_and_stochastic_for_snowdrift}
\end{figure}

Furthermore, to understand the impact of payoff memory on the snowdrift game
even more, we performed simulations for the four additional scenarios considered
in this work. As a reminder, in the first scenario, players recall the payoffs
they obtained in the last two rounds against a single co-player. In the second
scenario, players recall their last-round payoffs against two co-players. In the
third scenario, they recall the last two rounds against two co-players. Finally,
in the last scenario, players update based on the average payoff they receive
over all rounds with a single co-player (further extensions are possible, but we
do not explore them here). In Supplementary
Fig.~\ref{fig:cooperation_rate_all_updating_payoffs_snowdrift}, we show the
evolving cooperation rates for all these scenarios, as well as for the limited
and perfect memory cases.

We observe that main pattern of our results remains unchanged: the perfect
memory case always overestimates the cooperation rate. Compared to the baseline
case with a single round recalled, even a minimal increase in the players'
payoff memory further increases the cooperation rate. In all four scenarios with
extended memory, we observe similar cooperation rates, and they quickly approach
the rates observed under perfect memory. This happens at a faster rate compared
to the donation game.

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{static/snowdrift_game_more_memory.pdf}
  \caption{{\bf Average cooperation rates in the snowdrift game for different payoff memories.}
  We vary how much information individuals take into account when updating their strategies.
  From left to right, we consider the following cases:
  ({\it i}) Updating occurs based on expected payoffs (perfect memory).
  ({\it ii}) It occurs based on the last round of one interaction (limited memory).
  ({\it iii}) Based on the last round of two interactions.
  ({\it iv}) Based on the last two rounds of one interaction.
  ({\it v}) Based on the last two rounds of two interactions.
  ({\it vi}) Based on the average payoff of one interaction.
  Simulations are run either for a comparably low benefit of cooperation~($b/c\!=\!2$), or for a high benefit~($b/c\!=\!10$). 
  We observe that perfect memory always yields the highest cooperation rate, and limited memory yields the lowest.
  However, in the snowdrift game, the difference between the two extreme scenarios is not large, as we have already discussed.
  When individuals take into account a bit more information—cases (iii) to (vi)—evolving
  cooperation rates quickly approach the optimum of perfect memory. This increase happens faster compared to the donation game.
  Simulations are run for $T\!=\!10^7$ time steps, using a cost $c\!=\!1$, a continuation probability of $\delta\!=\!0.999$ and a selection strength of $\beta\!=\!1$, in a population of size $N\!=\!100$.}
\label{fig:cooperation_rate_all_updating_payoffs_snowdrift}
\end{figure}

@@

{
{\setlength{\bibsep}{0\baselineskip}
\bibliographystyle{naturemag}
\bibliography{bibliography}
}}
